---
title: "Trading ETF's - The sequel, Daniel Emil Wiinberg S133232"
output:
  pdf_document:
    latex_engine: lualatex
  html_notebook: default
  html_document:
    df_print: paged
---


```{r}
D <- read.table("finans2_data.csv", header = TRUE, sep = ";")
```
# Introduction to the data

```{r}
head(D); tail(D)
```

The data contain 95 rows of data (excluding header), one for each ETF. Each row contain 3 columns for the 3 variables: Geo.mean, Volatility and maxTuw. These variables are calculated from the data used in the first project, representing the 454 weeks returns of each ETF.

* **Geo.mean** contains observations of the geometric average rate of return for the ETFs in the dataset.
* **Volatility** represents the weekly volatility which is the standard deviation of the ratio between the price of an ETF at the beginning of a week and the end. 
* **maxTuW** is Maximum Time under Water, and indicates the maximum number of weeks between two peak prices.

## Scatterplots

First we plot the volatility and maxTuW variables against the geometric mean in a scatterplot.

```{r}
# Scatterplot
plot(c(D$Geo.mean), c(D$Volatility), type="p", xlab="Geo.mean", 
     ylab="Volatility", col=c("green"), cex=1)

plot(c(D$Geo.mean), c(D$maxTuW), type="p", xlab="Geo.mean", 
     ylab="maxTuW", col=c("blue"), cex=1)
```

## Histograms

Next we plot the three variables in a histogram plot to easily the distribution of values. 

```{r}
# Histograms
hist(D$Geo.mean, xlab="Geo.mean", prob=TRUE, breaks = 30, col="red")
hist(D$Volatility, xlab="Volatility", prob=TRUE, breaks = 30, col="green")
hist(D$maxTuW, xlab="maxTuW", prob=TRUE, breaks = 30, col="blue")
```

## Summary table

We extract some general information about the variables using the summary R function and display it in a table as shown below. 

```{r}
# summary(D)
# sd(D$Geo.mean)
# sd(D$Volatility
# sd(D$maxTuW)
```

|            |  n |   mean  |    sd   |  median |    q1   |   q3   |
|:----------:|:--:|:-------:|:-------:|:-------:|:-------:|:------:|
|  Geo.mean  | 95 | 0.07690 | 0.08087 | 0.08274 | 0.02871 | 0.1344 |
| Volatility | 95 |  3.060  |  0.8790 |  3.026  |  2.588  |  3.675 |
|   maxTuW   | 95 |  307.3  |  42.77  |  324.0  |  309.0  |  327.0 |

############################################################################

\newpage
# Multiple linear regression

Our linear regression model with for the geometric mean with our two variables for volatility and maxTuW is:
$$Y_i = \beta_0+\beta_1x_{1,i} + \beta_2 x_{2,i} + \epsilon_i$$
Where $Y_i$ is the geometric mean, $x_1$ is Volatility and $x_2$ is maxTuW. 


Next we divide out data set into two pools, where we will use the bigger pool to create a prediction model, and then use the smaller test set to see how well our prediction model behaves. 

```{r}
# Subset containing only AGG, VAW, IWN and SPY (for validation)
D_test = subset(D, ETF %in% c("AGG","VAW","IWN","SPY"))
# Subset containing only the 91 remaining ETFs (for model estimation)
D_model = subset(D, !(ETF %in% c("AGG","VAW","IWN","SPY")))
```


Here we create our linear regression for the geometric mean with our two variables.

```{r}
# Estimate multiple linear regression model
fit = lm(Geo.mean ~ Volatility + maxTuW, data = D_model)

# Show parameter estimates etc.
summary(fit)
```

From the summary is can be seen that:

| Name              | Value      |
|-------------------|------------|
|     $\beta_0$     |  0.2528394 |
|     $\beta_1$     | -0.0351310 |
|     $\beta_2$     | -0.0002203 |
|  $\hat{\sigma}^2$ |   0.07582  |
| Degree of freedom |     88     |
|       $R^2$       |   0.1516   |

$\beta_0$ is the interception with $x = 0$ and $\beta_1$ and $\beta_2$ are the slopes of the regression line. Meaning the effect that $\beta_i$ has on $x_i$. 

Since $\beta_1$ and $\beta_2$ are both slightly negative it means that volatility and Maximum Time under Water has a negative influence on the average return. 

## Model validation

We start by plotting the observed geometric mean values against our fitted values. 

```{r}
# Plots for model validation
# Observations against fitted values
plot(fit$fitted.values, D_model$Geo.mean, xlab = "Fitted values",     
       ylab = "Geom. average rate of return")
```

Next we plot the fitted values and each of our explanatory variables against the residuals. We are looking systematic dependences between the values and the residuals. 

```{r}
# Residuals against each of the explanatory variables
plot(D_model$Volatility, fit$residuals,
     xlab = "Volatility", ylab = "Residuals")
plot(D_model$maxTuW, fit$residuals,
     xlab = "maxTuW", ylab = "Residuals")

# Residuals against fitted values
plot(fit$fitted.values, fit$residuals, xlab = "Fitted values", 
     ylab = "Residuals")
```

From the plots of the residuals against the explanatory variables we do not see a systematic dependence, which means that our model is sound. 

Next we check if the residuals are normal distributed by doing a qq-plot.

```{r}
# Normal QQ-plot of the residuals
qqnorm(fit$residuals, ylab = "Residuals", xlab = "Z-scores", 
       main = "")
qqline(fit$residuals)
```

From the qq-plot and qq-line we can see that the residuals are normal distributed, which means that the normality assumption holds. 


#####################################################

\newpage
# Confidence interval for $\beta_1$

The confidence interval for parameter $\beta_1$ is given by:
$$\hat \beta_1 \pm t_{1-\alpha/2}\hat \sigma_{\beta_1}$$

```{r}
#Confidence interval for beta_1
-0.0351310 + c(1,-1)*qt(0.975, df=88)*0.0097229

# Confidence intervals for the model coefficients
confint(fit, level = 0.95)
```

We can see that the values calculated for $\beta_1$ using the formula are the same as when using the R-function. The confidence intervals for $\beta_0$ and $\beta_2$ are also given from the R-function above. 

## Hypothesis testing

We are testing whether $\beta_1$ might be -0.06, meaning we are testing: 
$$H_{0,1}: \beta_1 = \beta_{0,1} \Rightarrow \beta_1 = -0.06$$
Against the other hypothesis:
$$H_{1,1}: \beta_1 \neq \beta_{0,1} \Rightarrow \beta_1 \neq -0.06$$ 
We are testing with a significance level of $\alpha = 0.05$.
```{r}
tobs = (-0.0351310 - (-0.06)) / 0.0097229
tobs

# The p value becomes
2*(1-pt(tobs,df=88))
```

We can see that with a $p-value = 0.0122$ we have some evidence against the hypothesis. This could also be seen from the confidence interval for $\beta_1$ being $\{-0.0545; -0.0158\}$, since $-0.06$ is outside the interval.

#####################################################

\newpage
# Backward selection

If we go back to inspect our model, we can see that that maxTuW is not a significant parameter in our model, since its p-value is 0.252. Using backwards selection we remove this parameter from our model, and fit the data again.

```{r}
# Estimate multiple linear regression model
final_model = lm(Geo.mean ~ Volatility, data = D_model)

# Show parameter estimates etc.
summary(final_model)
```

We can now see that the slope is highly significant with a p-value of $9.62*10^{-5}$, meaning we cannot reduce our model further. 

Therefore our final model with estimations of our parameters is:
$$Y_i = \beta_0+\beta_1x_{1,i} = 0.19486 - 0.03824x_{1}$$

############################################################################

\newpage
# Comparing $D_{model}$ and $D_{test}$

First we predict the data for the geometric mean using our final. 

```{r}
# Predictions and 95% prediction intervals
pred = predict(final_model, newdata = D_test, 
               interval = "prediction", level = 0.95)
```

Next we use the cbind function to easily compare the actual geometric mean from our 4 test models to the values that we just got from our prediction. 

```{r}
# Observed values and predictions
compare = cbind(id = D_test$ETF, Geo.mean = D_test$Geo.mean, pred)
compare
```

We can see that the predictions are all fairly close, even though some of the predictions are better than others. However they are all within the 95% confidence interval. 

